Semantic Linearity in a Lazy Optimising Compiler

### 2024-01-29 icfp.tex
passed up to section 3.6.2
worked up to section 4

\subsubsection{Proof irrelevant resources}

@@

Resources used in a scrutinee that is not in weak head normal form must
definitely not be used in the case alternatives
% since they have been used in the evaluation of the scrutinee, as shown in
% the example above.
%
However, it is not sufficient to evaluate the scrutinee to weak head normal
form to \emph{fully} consume all resources used in the scrutinee, since
sub-expressions such as constructor arguments will be left unevaluated. To
\emph{fully} consume all resources occurring in the scrutinee, the scrutinee
must be evaluated either to \emph{normal form} or s.t. all linear components of
the scrutinee are fully evaluated, as witnessed by the $Alt0$ rule. In short,
for a case expression whose scrutinee is not in WHNF:
% We tackle this in due time, in the proof irrelevance section.
\begin{itemize}

\item The scrutinee resources must \emph{not} be used directly in the case alternatives;
\item But the result of evaluating the scrutinee to WHNF must still be
consumed, as all sub-expressions of the scrutinee remain unevaluated and must
be consumed.
\item Since the scrutinee resources cannot be consumed directly, they must be
consumed indirectly through $\D$-variables, namely, either the case binder, or
the linear pattern-bound variables introduced in the alternative.

\end{itemize}

@@

We introduce \emph{proof irrelevant} resources, denoted as linear resources
within square brackets $[\D]$, to encode linear resources that cannot be
directly used (the $Var$ rule is not applicable). Proof irrelevant resources
are linear resources in all other senses, meaning they must be used
\emph{exactly once}. However, since proof irrelevant resources cannot be
forgotten neither used directly, they have to be consumed \emph{indirectly} --
by $\D$-bound variables.

To type a case expression whose scrutinee is in weak head normal form, we
type the scrutinee with linear resources $\D$ and type the case
alternatives by introducing the case binder with a usage environment $[\D]$,
having the same proof irrelevant linear context $[\D]$ in the typing
environment, annotating the judgement with the proof irrelevant resources,
and using the $\Rrightarrow$ judgement:
\[
\TypeCaseNotWHNF
\]
Note how the rule is quite similar to the one for scrutinees in WHNF, only
diverging in that the resources in the case binder, typing environment, and
judgement annotation, are made irrelevant.

@@

Finally, we recall the tentative $Case_\textrm{WHNF}$ rule presented before and
highlight its flaw: the $\G;\D \vdash_{alt} \rho \to e :^z_{\D_s} \s \Mapsto
\vp$ judgement is only well-defined for patterns $\rho$ matching the WHNF form
of the scrutinee, as the distribution of resources per constructor components
only makes sense for the constructor pattern matching the scrutinee.
Essentially, if the scrutinee is in WHNF the matching alternative is easily
determined, and must be treated with the specialized $\Mapsto$ judgement, which
only applies to the matching constructor.
%
Alternatives not matching the scrutinee, as mentioned in the discussion of the
$Alt0$ rule, could use resources arbitrarily as they will never be executed,
however, we uniformly treat non-matching alternatives as if the scrutinee were
not in WHNF. The rule for typing case alternatives whose scrutinee is in WHNF
is thus given by:
\[
\TypeCaseWHNF
\]
We note that it might seem unusual to specialize a rule for expressions in
WHNF, as programs scrutinizing an expression in WHNF are rarely written by a
developer. Yet, our system is designed to be suitable for optimising compilers
in which intermediate programs commonly scrutinize expressions in WHNF.
%
Foreshadowing, type preservation for the case-reduction substituting the case
binder and pattern variables in the alternative is not possible to prove
without branching on WHNF-ness, since otherwise the $\D$-substitution is not
well-defined.

@@

Intuitively, in case alternatives whose scrutinee is not in weak head normal form,
% (and for scrutinees in WHNF which don't match the case alternative)
the proof-irrelevant resources introduced by the case expression must be fully
consumed, either via the case binder $z$, or by using all linear pattern-bound
variables (for uniformity, we also treat alternatives that do not match a
scrutinee in WHNF this way).

However, unlike with scrutinees in WHNF, the resources used by a
scrutinee not in WHNF do not necessarily match those used by each
sub-expression of the expression evaluated to WHNF.
%
Therefore, there is no direct mapping between the usage environments of the
linear pattern-bound variables and the resources used in the scrutinee.
% \todo{WAIT, there never was!
% Only when the K matches the scrutinee. Maybe we need a separate judgement for
% typing other constructors, which would be standalone and show up in this
% section too?}

We introduce \emph{tagged resources} to guarantee all linearly-bound pattern
variables are jointly used to consume all resources occurring in the
environment (in alternative to the case binder), or not at all. Given linear
resources $[\D_s]$ used to type a scrutinee, and a pattern
$K~\ov{x_\omega},\ov{y_i}$ with $i$ linear components, we assign a usage
environment $\D_i$ to each linear pattern variable where, $\D_i$ is obtained from the
scrutinee environment tagged with the constructor name and linear-variable
index $\lctag{\D_s}{K_i}$, and $\y[\D_i]$ is introduced in $\G$.
\[
  \begin{array}{cc}
    \TypeAltNNotWHNF & \TypeVarSplit
    \end{array}
\]
The tag consists of a constructor name $K$ and an index $i$ identifying the
position of the pattern variable among all bound variables in that pattern.
%
The key idea is that a linear resource $x$ can be split into $n$ resources at a
given constructor, where $n$ is the number of positional linear arguments of
the constructor.
%
This is given by the Split rule. By assigning to each linear pattern variable a fragment of the scrutinee
resources with a tag, we guarantee that all linear pattern variables are
simultaneously used to consume all the scrutinee resources, since for any of
scrutinee resources to be used by a linear pattern-bound var be used, the
resources must be $Split$ for the fragments corresponding to that $\D$-var to
be consumed, and, consequently, the remaining fragments have to be consumed
through the other linear pattern-bound variables.
%
For instance, in $\lambda x.~\ccase{x}{z~\{K~a~b\to (a,b)\}}$, where $x$ is a
linear variable, the case alternatives are typed with $\irr{x}$ (proof
irrelevant $x$), $z$ is introduced as $\z[\irr{x}]$, and the pattern variables
are introduced as $\var[a][\lctag{\irr{x}}{K_1}]$ and
$\var[b][\lctag{\irr{x}}{K_2}]$, assuming both components of $K$ are linear.
%
We note how $Split$ can be applied both to relevant and proof irrelevant linear resources in $\D$.

>>>>

\todo[inline]{Considerar alguns exemplos aqui para ilustrar esta
  ``análise de casos''}

\subsubsection{Pattern matching on non-WHNF terms via proof irrelevant resources}

@@

\todo[inline]{Todo este paragrafo parece-me redundante}
Resources used in a scrutinee that is not in WHNF must
not be used in the case alternatives.
%
%
However, it is not sufficient to evaluate the scrutinee to weak head normal
form to \emph{fully} consume all resources used in the scrutinee, since
sub-expressions such as constructor arguments will be left unevaluated. To
\emph{fully} consume all resources occurring in the scrutinee, the scrutinee
must be evaluated either to \emph{normal form} or such that all linear components of
the scrutinee are fully evaluated, as witnessed by the $Alt0$ rule above. In short,
for a case expression whose scrutinee is not in WHNF, the scrutinee
resources must \emph{not} be used directly in the case alternatives,
but the result of evaluating the scrutinee to WHNF must still be
consumed, as all sub-expressions of the scrutinee remain unevaluated and must
be consumed. Since the scrutinee resources cannot be consumed directly, they must be
consumed indirectly through $\D$-variables, namely, either the case binder, or
the linear pattern-bound variables introduced in the alternative.
% We tackle this in due time, in the proof irrelevance section.
% \begin{itemize}

% \item The scrutinee resources must \emph{not} be used directly in the case alternatives;
% \item But the result of evaluating the scrutinee to WHNF must still be
% consumed, as all sub-expressions of the scrutinee remain unevaluated and must
% be consumed.
% \item Since the scrutinee resources cannot be consumed directly, they must be
% consumed indirectly through $\D$-variables, namely, either the case binder, or
% the linear pattern-bound variables introduced in the alternative.

% \end{itemize}

@@

We introduce \emph{proof irrelevant} resources, written as linear resources
within square brackets $[ x{:}\s]$ and lifted to contexts as $[\D]$,
to encode linear resources that cannot be
directly used (i.e.,~to which the $Var$ rule is not applicable). Proof irrelevant resources
are linear resources in all other senses, meaning they must be used
\emph{exactly once}. However, since proof irrelevant resources cannot be
discarded nor used directly, they have to be consumed \emph{indirectly} --
by $\D$-bound variables.

To type a case expression whose scrutinee is in weak head normal form, we
type the scrutinee with linear resources $\D$ and type the case
alternatives by introducing the case binder with a usage environment $[\D]$,
having the same proof irrelevant linear context $[\D]$ in the typing
environment, annotating the judgement with the proof irrelevant resources,
and using the $\Rrightarrow$ judgement:
\[
\TypeCaseNotWHNF
\]
Note how the rule is similar to the one for scrutinees in WHNF, but
requiring the resources in the case binder, typing environment, and
judgement annotation to be made irrelevant.

@@

Finally, we recall the tentative $Case_\textrm{WHNF}$ rule presented before and
highlight its flaw: the $\G;\D \vdash_{alt} \rho \to e :^z_{\D_s} \s \Mapsto
\vp$ judgement is only well-defined for patterns $\rho$ matching the WHNF form
of the scrutinee, as the distribution of resources per constructor components
only makes sense for the constructor pattern matching the scrutinee.
% Essentially, if the scrutinee is in WHNF the matching alternative is easily
% determined, and must be treated with the specialized $\Mapsto$ judgement, which
% only applies to the matching constructor.
% %
Alternatives not matching the scrutinee, as mentioned in the discussion of the
$Alt0$ rule, could use resources arbitrarily as they will never be executed,
however, we uniformly treat non-matching alternatives as if the scrutinee were
not in WHNF. Having introduced proof irrelevant resources, we can now
present the updated and fully general $Case_\textrm{WHNF}$ rule:
\[
\TypeCaseWHNF
\]
We note that it might seem unusual to specialize a rule for expressions in
WHNF, as programs scrutinizing an expression in WHNF are rarely
written by humans.
Yet, our system is designed to be suitable for optimising compilers
in which intermediate programs commonly scrutinize expressions in WHNF.
%
Crucially, type preservation for the case-reduction substituting the
case binder and pattern variables in the alternative is not possible
to prove without branching on WHNF-ness, since otherwise the
$\D$-substitution is not well-defined.  \todo[inline]{What is the
  $\Delta$-substitution? Clarify}

@@

%% NOTA -- Sempre a repetir a mesma coisa
% Intuitively, in case alternatives whose scrutinee is not in WHNF,
% % (and for scrutinees in WHNF which don't match the case alternative)
% the proof-irrelevant resources introduced by the case expression must be fully
% consumed, either via the case binder $z$, or by using all linear pattern-bound
% variables (for uniformity, we also treat alternatives that do not match a
% scrutinee in WHNF this way).

Unlike with scrutinees in WHNF, the resources used by a
scrutinee not in WHNF do not necessarily match those used by each
sub-expression of the expression evaluated to WHNF.
%
Therefore, there is no direct mapping between the usage environments of the
linear pattern-bound variables and the resources used in the scrutinee.
% \todo{WAIT, there never was!
% Only when the K matches the scrutinee. Maybe we need a separate judgement for
% typing other constructors, which would be standalone and show up in this
% section too?}
%
We introduce \emph{tagged resources} to guarantee all linearly-bound pattern
variables are jointly used to consume all resources occurring in the
environment (in alternative to the case binder), or not at all. Given linear
resources $[\D_s]$ used to type a scrutinee, and a pattern
$K~\ov{x_\omega},\ov{y_i}$ with $i$ linear components, we assign a usage
environment $\D_i$ to each linear pattern variable where, $\D_i$ is obtained from the
scrutinee environment \emph{tagged} with the constructor name and linear-variable
index $\lctag{\D_s}{K_i}$, and $\y[\D_i]$ is introduced in $\G$.
\[
  \begin{array}{cc}
    \TypeAltNNotWHNF & \TypeVarSplit
    \end{array}
\]
The tag consists of a constructor name $K$ and an index $i$ identifying the
position of the pattern variable among all bound variables in that pattern.
%
The key idea is that a linear resource $x$ can be split into $n$ resources at a
given constructor, where $n$ is the number of positional linear arguments of
the constructor.
%
This is given by the Split rule. By assigning to each linear pattern
variable a fragment of the scrutinee resources with a tag, we
guarantee that all linear pattern variables are simultaneously used to
consume all the scrutinee resources.
For instance, in the term $\lambda x.~\ccase{x}{z~\{K~a~b\to
  (a,b)\}}$, where $x$ is a linear variable, the case alternative is
typed with $\irr{x}$,
the case binder $z$ is introduced as $\z[\irr{x}]$, and the pattern variables
are introduced as $\var[a][\lctag{\irr{x}}{K_1}]$ and
$\var[b][\lctag{\irr{x}}{K_2}]$, assuming both components of $K$ are linear.
\todo[inline]{Previous explanation was very confusing, try to explain
  what the split is doing in terms of the example!}
% since for any of the scrutinee
% resources to be used by a linear pattern-bound variable usage, the
% resources must be $Split$ for the fragments corresponding to that
% $\D$-var to be consumed, and, consequently, the remaining fragments
% have to be consumed through the other linear pattern-bound variables.

%
We note how $Split$ can be applied both to relevant and proof irrelevant linear resources in $\D$.
